---
title: "R Coding Exercise"
---

```{r}
#load package
library(dslabs)
library(dplyr)
library(ggplot2)
```

# review data

```{r}
#help for the gapminder data
help(gapminder)

#examine data structure
str(gapminder)

#summary of data
summary(gapminder)

#examine class
class(gapminder)

```

# generate new objects

```{r}
# Create object with African countries only
africadata <- gapminder %>%
  filter(continent == "Africa")

# Create a new object with life expectancy and infant mortality from African data
africa_mort_expect <- africadata %>%
  select(infant_mortality, life_expectancy)
	View(africa_mort_expect)

# View structure and summary of the mortality and life expectancy data
str(africa_mort_expect)
summary(africa_mort_expect)

# Create a new object with population and life expectancy from African data
africa_pop_expect <- africadata %>%
  select(population, life_expectancy)
View(africa_pop_expect)

# View structure and summary of the population and life expectancy data
str(africa_pop_expect)
summary(africa_pop_expect)

```

# create plots

## When plotted, there are distinct "streaks" in the data. This is especially evident in the Life Expectancy vs log$_{10}$ Population, These streaks likely represent the trajectory of individual countries over time. In general, as levels of development improve, so do health outcomes like life expectancy.

### Note: in both plots there is an outlier showing a visible decrease in life expectancy. Without countries or years in the plots, it is difficult to pin point the exact cause of that decrease, but it is likely due to a single, tranformational event. In this case, I would attribute that to the civil war and genocide in Rwanda in the early 1990s.

```{r}
library(ggplot2)

# Plot life expectancy as a function of infant mortality
plot_le_vs_im <- ggplot(africa_mort_expect, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Life Expectancy vs Infant Mortality",
       x = "Infant Mortality",
       y = "Life Expectancy")

plot_le_vs_im

#plot life expectancy vs population with population on a log scale
plot_le_vs_pop_log <- ggplot(africa_pop_expect, aes(x = population, y = life_expectancy)) +
    geom_point() +
    theme_minimal() +
    labs(title = "Life Expectancy vs Population",
         x = "Population (Log Scale)",
         y = "Life Expectancy") +
    scale_x_log10()

plot_le_vs_pop_log
```
